% !TEX root =  main.tex
\section{Introduction}
\label{sec:intro}

\subsection{Recursion and Dependent Types}
Dependently typed programming languages
 bridge the gap between theorem proving and programming.
In languages like Agda~\citep{agdaPaper}, Coq~\citep{coqart},
Idris~\citep{DBLP:journals/corr/abs-2104-00480}, and Lean~\citep{10.1007/978-3-319-21401-6_26},
 one can write programs, specifications, and proofs that programs
meet those specifications, all using a unified language.

One challenge in writing dependently typed code is proving termination.
Functions defined in dependently typed languages are typically required to be
\textit{total}: they must provably halt in all inputs.
This is necessary both to ensure that type checking terminates and to prevent
false results from being accidentally proven.
Since the halting problem
is undecidable, recursively-defined functions must be written in such a way that the type checker
can mechanically deduce termination.
Some functions only make recursive calls to structurally-smaller arguments,
so their termination is apparent to the compiler. However, some functions
cannot be easily expressed using structural recursion.
For such functions, the programmer must instead use \textit{well-founded recursion}, showing that there is some ordering, with no infinitely-descending
chains, for which each recursive call is strictly smaller according to this ordering. For example, the typical quicksort algorithm is not structurally recursive, but can use well-founded recursion on the length of the lists being sorted.

\subsection{Ordinals}

While numeric orderings work for first-order data, they are ill-suited to recursion over
higher-order data structures, where some fields contain functions.
Instead, one must use \textit{ordinals} to assign a size to such data structures, so that even
when a structure represents infinite data, only a finite number of recursive calls are made when traversing it.
In classical mathematics, ordinals are totally ordered and straightforward to reason about. They have
many different representations, all of which are equivalent.
However, in constructive theories, such as those underlying dependently typed languages,
there are many representations of ordinals which are not equivalent.
Different constructive ordinal notations have different capabilities, each with their own advantages and disadvantages.

\subsection{Contributions}

This work defines \textit{strictly monotone Brouwer Trees}, henceforth SMB-trees,
a new presentation of ordinals that hit a sort of sweet-spot for defining functions by
well-founded recursion. Specifically, SMB-trees:

\begin{itemize}
  \item Are strictly ordered by a well-founded relation;
  \item Have a maximum operator which computes a least-upper bound;
  \item Are \textit{strictly-monotone} with respect to the maximum: if $a < b$ and $c < d$, then $\max\ a\ c < \max\ b\ d$;
  \item Can compute the limits of arbitrary sequences;
  \item Are light in axiomatic requirements: they are defined without using axiom K,
        univalence, quotient types, or higher inductive types.
\end{itemize}

The novel insight behind our contribution is that there is a subset of
Brouwer trees which behave in the way we want. Specifically,
the ability of Brouwer trees to take the limit of a sequence allows
us to apply operations to an ordinal an infinite number of times,
exposing properties that do not hold for finite applications but do hold
in the limit.

\subsection{Uses for SMB-trees}

\subsubsection{Well-founded Recursion}

Having a maximum operator for ordinals is particularly useful when traversing over multiple higher order
data structures in parallel, where neither argument takes priority over the other.
In such a case, a lexicographic ordering cannot be used.

As an example, consider a unification algorithm that merges
two higher order data structures, such as a unifier for a strongly typed encoding of dependent types,
 and suppose that $\alpha$-renaming or some other restriction prevents
 structural recursion from being used.

To solve a unification problem $ \Sigma(x : A)\ldotp B = \Sigma(x : C)\ldotp D$
we must recursively
solve $A = C$ and $\forall x \ldotp B[x] = D[x]$.
However, the types of the variables in the latter equation are different.
So after computing the unification of $A$ and $C$, we may need to traverse
$B$ and $D$ and convert terms from type $A$ or $C$ to their unification.
If such a conversion is defined mutually with unification, then
it must work on a pair of types strictly smaller than
$\Sigma(x : A)\ldotp B , \Sigma(x : C)\ldotp D$.

To assign sizes to such a procedure, we need a few features.
First, we need
a maximum operator, so that we can bound the size of unifying $A$ and $C$
by their maximum size.
Second, the operator should be strictly monotone, so that the recursive
call unifying $A$ and $C$ is on a strictly smaller size.
Third, the maximum should be commutative:
we need the size of the nested pairs $((A,B),(C,D))$
to be the same as $((A,C),(B,D))$, so that a recursive call on arguments whose
size is bounded by the maximum of  $(A,C)$ will still
be strictly smaller than the initial size of $((A,B),(C,D))$.
One such call would be the procedure converting from type $A$ to the
solution of $A=C$.
Lexicographic orderings lack this commutativity, and are too restrictive
for situations such as this.

This style of well-founded induction was used to prove termination
in a syntactic model of gradual dependent types~\citep{Eremondi_2023}. There, Brouwer trees
were used to establish termination of recursive procedures that
combined the type information in two imprecise types.
The decreasing metric was the maximum size of the codes for the types being combined. Brouwer trees' arbitrary limits were used to assign sizes
to dependent function and product types, and the strict monotonicity of the
maximum operator was essential for proving that recursive calls were on
strictly smaller arguments.

In general, we want to provide the programmer with the ability
to specify complex relationships between the sizes of multiple arguments,
and to be able to deduce facts about those sizes in a principled way.

\subsubsection{Syntactic Models and Sized Types}

An alternate way view of our contribution is as a tool for modelling sized types~\citep{10.1145/237721.240882}.
The implementation of sized types in Agda has been shown to be unsound~\citep{agdaSizedIssue}, due to the interaction
between propositional equality and the top size $\infty$ satisfying $\infty < \infty$.
\Citet{Chan2022} defines a dependently typed language with sized types that does not have a top size, proving it consistent
using a syntactic model based on Brouwer trees.

SMB-trees provide the capability to extend existing syntactic models to sized types
with a maximum operator.
This brings the capability of consistent sized types closer to feature parity with Agda,
which has a maximum operator for its sizes,
while still maintaining logical consistency.

\subsubsection{Algebraic Reasoning}
Another advantage of SMB-trees is that they allow Brouwer trees to
be understood using algebraic tools.
In algebraic terminology, SMB-trees satisfy the following algebraic laws, up to the equivalence relation defined by $s \approx t := s \le t \le s $
\begin{itemize}
  \item Join-semlattice: the binary $\max$ is associative, commutative, and idempotent;
  \item Bounded: there is a least tree $Z$ such that $\max\ t\ Z \approx t$;
  \item Inflationary endomorphism: there is a successor operator $\up$
        such that ${\max\ (\up t)\ t \approx \up t}$
        and\\ ${\up (\max\ s\ t) \approx \max (\up s)\ (\up t)}$;
\end{itemize}

\Citet{BEZEM20221} describe a polynomial time algorithm for solving equations in such an algebra,
and describe its usefulness for solving constraints involving universe levels
in dependent type checking. While equations involving limits of infinite sequences
are undecidable, the inflationary laws could be used to automatically discharge some equations involving sizes. This algebraic presentation is particularly
amenable to solving equations using free extensions of algebras~\citep{corbyn:proof-synthesis,allais2023frex}.

% Even without sizes integrated into types, SMB-trees are useful for creating syntactic models of dependently typed languages and defining terms by well-founded recursion in these models.
% The ability to take the limit of arbitrary sequences of types makes it very easy to assign an ordinal size
% to encodings of dependent function or pair types: for $\Pi(x : A)\ldotp B$, if  $A$ has size $t_{A}$,
% and $B[x]$ has size $t_{B}[x]$ for each $x$, then the function type has size (using our notation from \cref{TODO})
% $\up (\max\ t_{A} (Lim\ A (\lambda x \ldotp B[x]))$:
% strictly larger than both the size of $A$ and the
% limit of size of $B$ for any $x : A$.
% The strict monotonicity property of SMB-trees let us compare function types:
% if $A_{1}$ is strictly smaller than $A_{2}$, and for every $x$ $B_{1}[x]$ is strictly smaller than $B_{2}[y]$ for some $y$, then strict monotonicity guarantees that $\Pi(x:A_{1})\ldotp B_{1}[x]$
% is strictly smaller than $$

\subsection{Implementation}

We have implemented SMB-trees in Agda 2.6.4.
Our library specifically avoids Agda-specific features
such as cubical type theory or Axiom K, so we expect
that the library can be easily ported to other proof assistants.

This paper is written as a literate Agda document, and the definitions
given in the paper are valid Agda code.
For several definitions, only the type is presented, with the body omitted due to
space restrictions. The full implementation can be found in the supplementary
materials section of this submission.
